The Impact of Alternation

### 2017-05-10 draft.tex
ok

  (\ref{it2:progress}) If $\tuple{\top,I_1,\dots,I_{n+1},\bot}$ is an
    interpolant for $\pi(u)$, the following entailments hold: \begin{compactitem}
    \item $\theta_0 \models I_1[Q_0/Q]$, 
      %
    \item $I_k[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k+1}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
      %
    \item $I_{n+1}[Q_{n+1}/Q] \wedge \theta_{n+1} \models \bot$. 
    \end{compactitem}
    We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n+1}$
    by induction on $n \geq 0$. This is sufficient to conclude because
    $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
    a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n+1} \wedge
    \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we
    have $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv
    \theta_0[Q/Q_0] \models I_1$. For the induction step $n>0$, we
    compute:
    \[\begin{array}{l}
    \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by definition of $\AbsPost{\A}$)} \\
    \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ ($\Asterisk$)} \\
    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q]  \hfill\text{\qed}
    \end{array}\]
   Let us prove the final point ($\Asterisk$). If
   $\I,\beta_n\cup\nu_{n-1}\cup\nu_n \models
   \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_{n}/Q]$,
   for some valuations $\beta_{n} : Q_{n-1} \rightarrow \booli$ and
   $\nu_i : \vec{x}_i \rightarrow \Data^\I$, for $i=n-1,n$, then we
   build a valuation $\beta_{n-1}$ such that
   $\I,\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
   \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$,
   as follows. For each occurrence of a subformula $\Delta(q,a_n)$ in
   $\AbsPost{\A}(\iota,a_1\ldots a_{n-1})$ we set
   $\beta_{n-1}(q_{n-1}) = \true$ if $\I,\beta_n\cup\nu_{n-1}\cup\nu_n
   \models \Delta(q,a_n)[Q_n/Q]$ and $\beta_{n-1}(q_{n-1}) = \false$,
   otherwise. Since there are no negated occurrences of such
   subformulae, the definition of $\beta_{n-1}$ is consistent, and the
   check $\I,\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
   \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$
   is straightforward. \qed}

@@

  \State let $\Art = \tuple{N,E,\Lambda,\lhd}$ be an ART

@@

  For the second point, we prove the following invariant: each time
  Algorithm \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$
  of nodes in $\worklist$ contains all the frontier nodes in $\Art =
  \tuple{N,E,\Lambda,\lhd}$ which are not covered by some node in $N$,
  namely that $W = \set{n \in N \mid \forall m \in N \forall a \in
    \Sigma ~.~ (n,a,m) \not\in E ~\wedge~ (n,m) \not\in
    \lhd}$. Initially, this is the case because $W = \set{\rootNode}$
  and $E = \lhd = \emptyset$. If the invariant holds previously, at
  line \ref{ln:while}, it will hold again after line \ref{ln:refine}
  is executed, because, when the subtree rooted at the pivot $p$ is
  removed, $p$ becomes a member of the set of uncovered frontier
  nodes, and is added to $W$ at line \ref{ln:refine}. The last case
  considers that the invariant holds at line \ref{ln:while} and the
  control follows the else branch at line \ref{ln:else}. In this case,
  the newly created frontier node $s$ is added to $W$ only if it is
  not covered by an existing node in $N$ (line \ref{ln:covered}).

@@

The ART built by Algorithm \ref{alg:impact} corresponds to an
unfolding of the input automaton by enumerating the input event
sequences in $\Sigma^*$. Each time a new node is created, its label is
initially $\top$. This label is subject to subsequent strenghtening,
when a spurious counterexample is encountered. To speedup the
computation of interpolants for spurious counterexamples, we use two
additional labeling functions for bookkeeping. Formally, the ART
structure used by Algorithm \ref{alg:impact} is defined as $\Art =
\tuple{N,E,\Lambda,R,T,\lhd}$, where $N,E,\Lambda$ and $\lhd$ are
defined as for Algorithm \ref{alg:predabs} and,
moreover: \begin{compactitem}
\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$, 
%
\item $T : E \rightarrow \bigcup_{i=0}^\infty
  \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is labeling of edges
  with time-stamped formulae, and
%
\item for each edge $e = (n,a,m) \in E$, if $n$ lies at
  depth\footnote{We consider that the root node has zero depth.}  $k$
  in the tree $(N,E)$, then we have:
  \[\begin{array}{rcl}
  T(e) & = & \Theta(n,k) \\
  R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)} \text{ where} \\
  \Theta(n,k) & \equiv & \bigwedge_{q \in R(n)} (q_k \rightarrow
  \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}])
  \end{array}\]
\end{compactitem}
For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is
$(n,a,m)$. This additional bookkeeping is used as follows. Given a
sequence $u = a_1\ldots a_k \in \Sigma^*$ that labels the path from
the root to a node $n$ in $\Art$, the conjunction:
\[\mu(n) \equiv \Lambda(\rootNode) \wedge T(\rootNode,a_1,n_1) \wedge \ldots
T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q
\rightarrow \bot)\] defines the interpolation problem $\pi(u)$.

@@

A node $n \in N$ is \emph{covered} if it has an ancestor $m$ such that
$(m,p) \in \lhd$, for some $p \in N$. A node $n$ is \emph{closed} if
it is covered, or $\Lambda(n) \not\models \Lambda(m)$ for all $m \in
N$ such that $n \prec^* m$.

@@

  \State let $\Art = \tuple{N,E,\Lambda,R,T,\lhd}$ be an ART

@@

  \While{$\worklist \neq \emptyset$}

@@

@@

  \State let $a_1 \ldots a_k$ be the label of the path from
  $\rootNode$ to $n$

@@

  \State let $\rootNode,n_1,\ldots,n_k,n_{k+1}=n$ be the sequence of nodes from $\rootNode$ to $n$
  \label{ln:refine-begin}

  \State let $\tuple{\top,I_1,\ldots,I_k,I_{k+1}\equiv\bot}$ be an interpolant for $\mu(n)$

@@

  \For{$i=1,\ldots,k+1$}

@@

  \State $\lhd \leftarrow \lhd \setminus \set{(m,n_i) \in \lhd \mid m \in N}$
  \Comment{uncover $n_i$}

@@

  \State $E \leftarrow E \cup \set{e}$

@@

\State $\lhd \leftarrow \left(\lhd \setminus \set{ (p,q) \in \lhd \mid
  \text{$q$ is a successor of $x$}}\right) \cup \set{(x,y)}$
\label{ln:close-uncover}

@@

The execution of Algorithm \ref{alg:impact} consists of three
phases\footnote{These phases correspond to the \textsc{Close},
  \textsc{Refine} and \textsc{Expand} procedures from the original
  description of \impact~\cite{mcmillan06}.}: \emph{close},
\emph{refine} and \emph{expand}. When a node $n$ is removed from the
worklist, it is first closed (line \ref{ln:close-worklist}) by finding
a lexicographically bigger node $m$ such that $\Lambda(n) \models
\Lambda(m)$, if one exists, and adding $(n,m)$ to $\lhd$. As in
Algorithm \ref{alg:predabs}, the intuition is that for any feasible
counterexample that can be discovered from $n$ there exists a feasible
counterexample stemming in $m$, thus we do not need to process $n$
further.

If $n$ has not been previously covered at line
\ref{ln:close-worklist}, we check if it is the endpoint of a
counterexample candidate, which can only be the case when the
replacement set $R(n)$ contains at least one final state (line
\ref{ln:cex}). Observe that, if $u$ is the label of the path from
$\rootNode$ to $n$, then $\Accept{\A}(u)$ is satisfiable only if it
contains a (positive occurrence of) a final state. The counterexample
is moreover feasible iff $\mu(n)$ is satisfiable, in which case a model of
$\mu(n)$ is obtained and a word $w \in L(\A)$ returned.

Otherwise, the counterexample is spurious and the procedure enters the
refinement phase (lines \ref{ln:refine-begin} -
\ref{ln:refine-end}). The interpolant for $\mu(n)$ is used to
strenghten the labels of all the ancestors of $n$, by conjoining the
formulae of the interpolant to the existing labels. In this process,
some node might become eligible for coverage, therefore we attempt to
close each ancestor of $n$ that is impacted by the refinement (line
\ref{ln:refine-end}). Observe that, in this case the call to
$\Call{Close}{}$ must uncover each node which is covered by a
successor of $n$ (line \ref{ln:close-uncover} in the $\Call{Close}{}$
procedure). This is required because, due to the over-approximation of
the sets of reachable configurations, the covering relation is not
transitive, as explained in \cite{mcmillan06}.

@@

>>>>

  (\ref{it2:progress}) If $\tuple{\top,I_1,\dots,I_{n+1},\bot}$ is an
    interpolant for $\pi(u)$, the following entailments hold: \begin{compactitem}
    \item $\theta_0 \models I_1[Q_0/Q]$, 
      %
    \item $I_k[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k+1}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
      %
    \item $I_{n+1}[Q_{n+1}/Q] \wedge \theta_{n+1} \models \bot$. 
    \end{compactitem}
    We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n+1}$
    by induction on $n \geq 0$. This is sufficient to conclude because
    $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
    a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n+1} \wedge
    \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we
    have $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv
    \theta_0[Q/Q_0] \models I_1$. For the induction step $n>0$, we
    compute:
    \[\begin{array}{l}
    \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by definition of $\AbsPost{\A}$)} \\
    \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ ($\Asterisk$)} \\
    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q] 
    \end{array}\]
   Let us prove the final point ($\Asterisk$). If
   $\I,\beta_n\cup\nu_{n-1}\cup\nu_n \models
   \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_{n}/Q]$,
   for some valuations $\beta_{n} : Q_{n-1} \rightarrow \booli$ and
   $\nu_i : \vec{x}_i \rightarrow \Data^\I$, for $i=n-1,n$, then we
   build a valuation $\beta_{n-1}$ such that
   $\I,\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
   \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$,
   as follows. For each occurrence of a subformula $\Delta(q,a_n)$ in
   $\AbsPost{\A}(\iota,a_1\ldots a_{n-1})$ we set
   $\beta_{n-1}(q_{n-1}) = \true$ if $\I,\beta_n\cup\nu_{n-1}\cup\nu_n
   \models \Delta(q,a_n)[Q_n/Q]$ and $\beta_{n-1}(q_{n-1}) = \false$,
   otherwise. Since there are no negated occurrences of such
   subformulae, the definition of $\beta_{n-1}$ is consistent, and the
   check $\I,\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
   \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$
   is straightforward. \qed}

@@

  \State let $\Art = \tuple{N,E,\Lambda,\lhd}$ be an ART with root
  node $\rootNode$

@@

  For the second point, we prove the following invariant: each time
  Algorithm \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$
  of nodes in $\worklist$ contains all the frontier nodes in the ART
  $\tuple{N\cup W,E,\Lambda,\lhd}$ which are not covered by some node
  in $N$, namely that:
  \begin{equation}\label{inv:predabs}
  W = \set{n \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
    \not\in E ~\wedge~ (n,m) \not\in \lhd}
  \end{equation}
  Initially, this is the case because $W = \set{\rootNode}$ and $N = E
  = \lhd = \emptyset$. If the invariant holds previously, at line
  \ref{ln:while}, it will hold again after line \ref{ln:refine} is
  executed, because, when the subtree rooted at the pivot $p$ is
  removed, $p$ becomes a member of the set of uncovered frontier
  nodes, and is added to $W$ at line \ref{ln:refine}. The last case
  considers that the invariant holds at line \ref{ln:while} and the
  control follows the else branch at line \ref{ln:else}. In this case,
  the newly created frontier node $s$ is added to $W$ only if it is
  not covered by an existing node in $N$ (line \ref{ln:covered}).

@@

The ART built by Algorithm \ref{alg:impact} corresponds to an
unfolding of the input automaton by enumerating the input event
sequences in $\Sigma^*$. Each time a new node is created, its label is
initially $\top$. This label is subject to subsequent strenghtening,
when a spurious counterexample is encountered. To speedup the
computation of interpolants for spurious counterexamples, we use two
additional labeling functions for bookkeeping. Formally, the ART
structure used by Algorithm \ref{alg:impact} is defined as $\Art =
\tuple{N,E,\Lambda,R,T,\lhd}$, where $N,E,\Lambda$ and $\lhd$ are
defined as for Algorithm \ref{alg:predabs} and,
moreover: \begin{compactitem}
\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$, 
%
\item $T : E \rightarrow \bigcup_{i=0}^\infty
  \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is labeling of edges
  with time-stamped formulae, and
%
\item for each edge $e = (n,a,m) \in E$, if $n$ lies at
  depth\footnote{We consider that the root node has zero depth.}  $k$
  in the tree $(N,E)$, then we have:
  \[\begin{array}{rcl}
  T(e) & = & \Theta(n,k) \\
  R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)} \text{ where} \\
  \Theta(n,k) & \equiv & \bigwedge_{q \in R(n)} (q_k \rightarrow
  \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}])
  \end{array}\]
\end{compactitem}
For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is
$(n,a,m)$. This additional bookkeeping is used as follows. Given a
sequence $u = a_1\ldots a_k \in \Sigma^*$ that labels the path from
the root to a node $n$ in $\Art$, the conjunction:
\[\mu(n) \equiv \Lambda(\rootNode) \wedge T(\rootNode,a_1,n_1) \wedge \ldots
T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q_k
\rightarrow \bot)\] defines the interpolation problem $\pi(u)$.

@@

A node $n \in N$ is \emph{covered} if $(n,p) \in \lhd$ or it has an
ancestor $m$ such that $(m,p) \in \lhd$, for some $p \in N$. A node
$n$ is \emph{closed} if it is covered, or $\Lambda(n) \not\models
\Lambda(m)$ for all $m \in N$ such that $n \prec^* m$.

@@

  \State let $\Art = \tuple{N,E,\Lambda,R,T,\lhd}$ be an ART with root
  node $\rootNode$

@@

  \While{$\worklist \neq \emptyset$}
  \label{ln:impact-while}

@@

  \State let $(\rootNode,a_1,n_1),(n_1,a_2,n_2), \ldots,
  (n_{k-1},a_k,n)$ be the path from $\rootNode$ to $n$

@@

@@

  \State let $\tuple{\top,I_1,\ldots,I_k \equiv\bot}$ be an interpolant for $\mu(n)$
  \label{ln:refine-begin}

@@

  \For{$i=1,\ldots,k$}

@@

  \State $\lhd \leftarrow \lhd \setminus \set{(m,n_i) \in \lhd \mid m \in N}$

@@

  \State $E \leftarrow E \cup \set{e}$
  \label{ln:edge-insert}

@@

\State $\lhd \leftarrow \left(\lhd \setminus \set{ (p,q) \in \lhd \mid
  \text{$q$ is $x$ or a successor of $x$}}\right) \cup \set{(x,y)}$
\label{ln:close-uncover}

@@

The execution of Algorithm \ref{alg:impact} consists of three
phases\footnote{These phases correspond to the \textsc{Close},
  \textsc{Refine} and \textsc{Expand} procedures from the original
  description of \impact~\cite{mcmillan06}.}: \emph{close},
\emph{refine} and \emph{expand}. When a node $n$ is removed from the
worklist, it is first closed (line \ref{ln:close-worklist}) by finding
a lexicographically bigger node $m$ such that $\Lambda(n) \models
\Lambda(m)$, if one exists, and adding the pair $(n,m)$ to $\lhd$. As
in Algorithm \ref{alg:predabs}, the intuition is that for any feasible
counterexample that can be discovered from $n$ there exists a feasible
counterexample stemming in $m$, thus we do not need to process $n$
further.

If $n$ is not covered, we check if it is the endpoint of a
counterexample candidate, which can only be the case when the
replacement set $R(n)$ contains at least one final state (line
\ref{ln:cex}). Observe that, if $u$ is the label of the path from
$\rootNode$ to $n$, then $\Accept{\A}(u)$ is satisfiable only if it
contains a (positive occurrence of) a final state. The counterexample
is moreover feasible iff $\mu(n)$ is satisfiable, in which case a
model of $\mu(n)$ is obtained and a word $w \in L(\A)$ returned.

Otherwise, the counterexample is spurious and the procedure enters the
refinement phase (lines \ref{ln:refine-begin} -
\ref{ln:refine-end}). The interpolant for $\mu(n)$ is used to
strenghten the labels of all the ancestors of $n$, by conjoining the
formulae of the interpolant to the existing labels. Then $\Lambda(n)$
becomes equivalent to $\bot$.

In this process, the nodes on the path between $\rootNode$ and $n$,
including $n$, might become eligible for coverage, therefore we
attempt to close each ancestor of $n$ that is impacted by the
refinement (line \ref{ln:refine-end}). Observe that, in this case the
call to $\Call{Close}{}$ must uncover each node which is covered by a
successor of $n$ (line \ref{ln:close-uncover} in the $\Call{Close}{}$
procedure). This is required because, due to the over-approximation of
the sets of reachable configurations, the covering relation is not
transitive, as explained in \cite{mcmillan06}. If $\Call{Close}{}$
adds a covering edge $(n_i,m)$ to $\lhd$, it does not have to be
called for the successors of $n_i$ on this path, which is handled via
the boolean flag $b$.

@@

The correctness of Algorithm \ref{alg:impact} is proved in the
following:

\begin{theorem}\label{thm:impact}
  Given an automaton $\A$, such that $L(\A) \neq \emptyset$, Algorithm
  \ref{alg:impact} terminates and returns a word $w \in L(\A)$. If
  Algorithm \ref{alg:impact} terminates reporting $\true$, then
  $L(\A) = \emptyset$.
\end{theorem}
\proof{ The first point is proved along the lines of the first point
  of Theorem \ref{thm:predabs}. In particular, it is easy to see that
  each call to $\Call{Close}{}$ terminates, because $N$ is finite at
  each moment during the execution.

  For the second point, we prove the following invariant: each time
  Algorithm \ref{alg:impact} reaches the line \ref{ln:impact-while},
  the set of nodes $W$ in $\worklist$ contains exactly those nodes in
  the ART $\tuple{N \cup W,E,\Lambda,R,T,\lhd}$ which are
  both\begin{inparaenum}[(i)]
  \item uncovered, and such that
  \item for some $a \in \Sigma$, the edge $(n,a,s)$ is missing from $E$:
  \end{inparaenum}
  \begin{equation}\label{inv:impact}
  W = \set{n \mid n \text{ is uncovered and } \exists a \in \Sigma~
    \forall s \in N ~.~ (n,a,s) \not\in E}
  \end{equation}
  Initially, $W = \set{\rootNode}$ and $N=E=\lhd=\emptyset$, thus the
  invariant holds trivially. Suppose that the invariant holds at some
  point, when reaching line \ref{ln:impact-while} and some node $n$
  was removed from $W$ and inserted into $N$. We distinguish two
  cases, either: \begin{itemize}
  \item $n$ is covered at line \ref{ln:close-worklist}, in which case
    $W$ remains unchanged, or
  %
  \item $n$ is not covered, in which case for each $a \in \Sigma$ a
    new node $s$ is inserted into $W$ (line \ref{ln:expand-end}) and
    the edge $(n,a,s)$ is inserted into $E$ (line
    \ref{ln:edge-insert}). None of these new nodes are covered and
    none of their successors inserted into $E$.
  \end{itemize}
  Then the condition (\ref{inv:impact}) holds next time the line
  \ref{ln:impact-while} is reached, thus proving the invariant.  \qed}
